FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-27 manuscript.Rmd
Update manuscript.Rmd
first update

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely produce hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
as the theory's central claims evade falsification.
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
a philosophical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \@ref(fig:figecs)).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

```{r eval = FALSE}
library(tidySEM)
library(ggplot2)
y_spacing <- .5
lo <- get_layout("", "A", "",
                 "D", "", "B",
                 "", "C", "", rows = 3)

@@

```{r figec, eval = FALSE, fig.cap="A take on the empirical cycle by De Groot"}
knitr::include_graphics("empirical_cycle.pdf")
```

@@

We will use the empirical cycle as a running example for this tutorial.
The resulting FAIR version of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.
The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969,
<!-- Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking". -->
<!-- We can thus consider it a meta-theory of theory construction. -->
<!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
and consists of a series of natural language statements:

@@

}')
rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.svg")
rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
rsvg::rsvg_pdf(svg = "ec_groot.svg",
               file = "ec_groot.pdf")
rsvg::rsvg_pdf(svg = "ec_wagenmakers.svg",
               file = "ec_wagenmakers.pdf")
rsvg::rsvg_pdf(svg = "ec_lissa.svg",
               file = "ec_lissa.pdf")
```

```{r figecs, figures-side, fig.show="hold", out.width="33%", fig.cap="The empirical cycle according to De Groot (left), Wagenmakers and colleagues (center), and Van Lissa (right)"}
knitr::include_graphics("ec_groot.pdf")
knitr::include_graphics("ec_wagenmakers.pdf")
knitr::include_graphics("ec_lissa.pdf")
```

@@

## 2. Creating the Project Folder

We can write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`, and save it to a new folder.
To help meet the Interoperability and Reusability criteria,
add two more files.
Include a `README.md` file with instructions for future users of your theory (see: [What to Include in a README?]())
and a `LICENSE` file with the legal conditions for reuse.
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
To version control our project, we initiate a Git repository in the project folder.
<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
<!-- opening a terminal in the project folder, -->
<!-- and running the command `git init`. -->
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`. -->
<!-- The default settings for the repository are fine; -->
<!-- for the next steps of the tutorial it  -->
Note that the repository
must be set to "Public" to take advantage of GitHub's Zenodo integration.
<!-- Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`. -->
<!-- Next, in the terminal window, run the following lines of code sequentially: -->
<!-- ``` -->
<!-- git add . -->
<!-- git commit -m "First commit" -->
<!-- git remote add origin *[the repository URL you copied]* -->
<!-- git branch -M main -->
<!-- git push -u origin main -->
<!-- ``` -->
Push the local files to the Git remote repository, and keep them synchronized going forward.

>>>>

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to the widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked. <!-- NvD: In philosophy of science, induction is considered to be problematic (logically invalid). It is more appropriate to talk about abduction when it comes to theory development. I am fine with sticking to induction if that is what psychologists "know". In that case, I suggest adding a clarifying footnote on abduction vs induction.  -->
The present paper restores balance by applying, for the first time, <!-- NvD: Are we sure that this is the first time? Do we need to make this claim? -->
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the *empirical cycle* [@degrootMethodologieGrondslagenVan1961],
a meta-theoretical model of cumulative knowledge acquisition, <!-- NvD: De Groot was not a philosopher and his model is not commonly known/accepted in philosophy of science -->
research ideally follows a cyclical process with two phases (Figure \@ref(fig:figec)).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

```{r}
library(tidySEM)
library(ggplot2)
y_spacing <- .5
lo <- get_layout("", "A", "",
                 "D", "", "B",
                 "", "C", "", rows = 3)

@@

```{r figec, fig.cap="A take on the empirical cycle by De Groot"}
knitr::include_graphics("empirical_cycle.pdf")
```

@@

Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the *empirical cycle*,
we decided to use it as an example for this tutorial.
The resulting FAIR theory is available at <https://doi.org/10.5281/zenodo.14552329>.
The empirical cycle is described on page 28 of @degrootMethodologyFoundationsInference1969.
Note that, while De Groot does not explicitly refer to the empirical cycle as a "theory", he derives it from "a theory of thinking".
We can thus consider it a meta-theory of theory construction.
<!-- The first step in making a theory FAIR is constructing a specific *implementation* [@guestHowComputationalModeling2021]. -->
The "empirical cycle" theory presented by De Groot consists of a series of natural language statements:

@@

}')
rsvg_svg(charToRaw(export_svg(ec_groot)), "ec_groot.svg")
rsvg_svg(charToRaw(export_svg(ec_wagenmakers)), "ec_wagenmakers.svg")
rsvg_svg(charToRaw(export_svg(ec_lissa)), "ec_lissa.svg")
```

@@

We can now write this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.

## 2. Creating a Project Folder

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

### What's in a README?

The readme should contain information to help people get started with using your FAIR theory.
We suggest the following elements:

* Title, prefaced with `# FAIR theory: The Theory's Name`
* Description: A plain-text description of the theory and its scope
* Interoperability: Most README files contain a section labeled "Getting Started", "Instructions", or "How to Use". From a FAIR perspective, such a section might be better labeled "Interoperability", or "How to Use (Interoperability)". We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
* Contributing: Pertaining to the Reusability criterion, this section should tell users the *social expectations regarding reuse and contributions*.
* License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the *legal conditions of reuse*.
* Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
* Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here:

```
This repository contains an implementation of the "empirical cycle",
a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.

> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
Foundations of inference and research in the behavioral sciences.
De Gruyter Mouton. https://doi.org/10.1515/9783112313121
```

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
To version control our project, we initiate a Git repository in the project folder.
<!-- This can be done by [installing Git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) locally, -->
<!-- opening a terminal in the project folder, -->
<!-- and running the command `git init`. -->
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
<!-- To [do this on GitHub](https://github.blog/developer-skills/github/beginners-guide-to-github-repositories-how-to-create-your-first-repo/), create an account and press the button labeled `Create new repository`. -->
<!-- The default settings for the repository are fine; -->
<!-- for the next steps of the tutorial it  -->
Note that the repository
must be set to "Public" to take advantage of GitHub's Zenodo integration.
<!-- Copy the repository's URL, which usually has the form `https://github.com/username/repository.git`. -->
<!-- Next, in the terminal window, run the following lines of code sequentially: -->

<!-- ``` -->
<!-- git add . -->
<!-- git commit -m "First commit" -->
<!-- git remote add origin *[the repository URL you copied]* -->
<!-- git branch -M main -->
<!-- git push -u origin main -->
<!-- ``` -->
Push the local files to the Git remote repository, and keep them synchronized going forward.
